Beef Pricing Dynamics in the Canadian Grocery Market: A Comparative Analysis of Walmart and T&T

### 2024-11-28 paper/paper.qmd
all text revised according to rubric word list

The pricing of grocery products, particularly beef, plays a significant role in understanding consumer behavior and market dynamics within the Canadian retail sector. Beef, as a staple of many households, exhibits complex pricing behaviors influenced by a variety of factors, including vendor strategies [@citePricingStrategies], historical pricing, and seasonal trends [@citeSeasonal]. The broader context for this study lies in exploring how large-scale retailers and specialty stores adjust their beef prices over time and how these changes reflect broader economic and cultural trends. Given the increasing concerns around food affordability and accessibility, understanding the dynamics of beef pricing is not only crucial for consumers but also for stakeholders involved in food distribution and policy-making.

This paper specifically focuses on two prominent vendors, Walmart and T&T, to examine how beef pricing varies across different types of retail environments. Walmart, as a major multinational retailer, represents mainstream [@citeWalmart], standardized pricing strategies, while T&T, catering primarily to Asian communities, represents a more specialized market segment [@citeTandT]. By analyzing data from these two vendors, this paper aims to address an important gap in understanding how cultural preferences and retail strategies influence product pricing. While previous studies have highlighted general trends in food pricing, they often overlook the nuanced differences between general retail chains and culturally specific stores. This paper fills that gap by providing an in-depth analysis of beef pricing patterns across these two distinct retail settings.

@@

Using a Bayesian linear regression approach, this study examines the effect of historical pricing, vendor-specific strategies, and seasonal factors on current beef prices. The analysis finds that old prices significantly predict current pricing, indicating a high degree of price consistency. Additionally, notable differences in pricing strategies between Walmart and T&T suggest that these vendors approach their markets with distinct pricing objectives, with Walmart favoring price stability and T&T showing more variability in pricing, possibly to cater to niche preferences. The findings are significant as they reveal the underlying mechanisms of price determination in a diverse retail environment, highlighting both market-wide factors and vendor-specific influences. Such insights could be useful for retailers looking to optimize their pricing strategies and for policymakers interested in addressing food affordability concerns.

The structure of the paper is organized as follows: @sec-data outlines the data collection and cleaning process, along with a description of the outcome and predictor variables used in the analysis. @sec-model introduces the forecasting models and discusses the rationale behind choosing these models for beef pricing prediction. @sec-result then presents the main findings, including a breakdown of vendor-level pricing effects and seasonal trends. Finally, @sec-discussion interprets the results, highlighting significant trends, comparing vendor strategies, and discussing potential limitations of the analysis.

@@

The dataset used for this research was sourced from Hammer, a publicly available repository designed to provide pricing data from various retail chains. The dataset comprises detailed information on product attributes, such as product name, vendor, current price, old price, units available, and additional details like the month of certain price. The dataset's broader context is centered on consumer pricing trends across different vendors, helping to understand how factors like vendor type, previous pricing, and seasonal changes influence current prices. This information is pivotal for examining market behaviors and assessing pricing strategies in retail environments.

The dataset variables include several categorical and numerical components. Key variables such as ‘vendor’, which identifies the retailer (e.g., Walmart, Galleria), and ‘product_name’, which details the item, are crucial for understanding distribution channels. Variables like ‘current_price’ and ‘old_price’ offer insights into pricing dynamics, enabling analyses of price changes over time. Additionally, temporal variables such as ‘month’ allow for the identification of seasonal variations in pricing. Summary statistics, along with graphs for each of these variables, help illustrate their distributions and relationships. Graphs have been included to show the frequency distribution for vendors, trends in price changes, and comparisons between old and current prices. Relationships among variables like ‘vendor’ and ‘current_price’ have been highlighted to give a comprehensive view of the dataset.

There were other potential datasets available for this analysis, such as proprietary retail data sources or other consumer purchasing databases. However, they were not selected due to restrictions on data availability, licensing requirements, or lack of detail in specific product-level attributes. The Hammer dataset was chosen as it provides granular pricing data that is crucial for understanding product-level trends across multiple vendors. The dataset also allowed for the construction of new variables, such as the calculated ‘price_per_unit’, which enabled more meaningful comparisons between products. High-level data cleaning included handling missing values in the ‘old_price’ variable by replacing them with the ‘current_price’, ensuring the analysis was consistent and complete.

@@

The process of measurement in this research involves translating real-world phenomena into quantifiable data entries within the dataset. For instance, consumer purchasing behavior, influenced by factors such as vendor type, pricing history, and seasonal changes, has been captured through variables like ‘vendor’, ‘current_price’, and ‘month’. The ‘vendor’ variable serves as an identifier of the source of the product, which helps to provide insights into how different retail environments may affect pricing strategies. The ‘current_price’ and ‘old_price’ variables reflect pricing trends and allow us to measure the temporal changes in consumer costs, thereby linking economic activities to specific data points.

The temporal dimension, captured through the ‘month’ variable, enables us to measure the impact of seasonal patterns on pricing. Constructed variables such as ‘price_per_unit’ provide a standardized measurement that allows comparisons across different products, accounting for varying package sizes or quantities. This transformation from abstract consumer behaviors and retail dynamics into structured data ensures that the analysis remains rooted in real-world market phenomena, thereby enabling a more nuanced understanding of pricing behavior and trends across different vendors.

## Outcome variables
The outcome variable for this study, current_price, serves as the key focus of analysis in understanding product pricing dynamics. This section provides a structured overview of the current_price variable, including its distribution and any observed trends, supported by both summary statistics and visualizations. The aim is to offer a clear understanding of how product prices vary and what factors may influence those variations.

### Current Price
The outcome variable in this study is current_price. This variable represents the price of the product at the time of data collection and is central to understanding the effects of various factors on product pricing. By modeling current_price, we aim to explore how factors such as historical pricing, vendor type, and seasonal influences impact current market prices. This variable is crucial for determining the pricing dynamics employed by different vendors and how those strategies affect consumer costs.

To provide a comprehensive view of the outcome variable, Here's a table for easier understanding (@tbl-current_price). This table provides the minimum, maximum, mean, median, and standard deviation of the current_price variable, offering insights into the overall distribution and variability in product prices.

@@

outcome_summary <- beef_data %>%
  summarise(
    Statistic = c("Min", "Max", "Mean", "Median", "Standard Deviation"),
    Value = c(
      min(current_price, na.rm = TRUE),
      max(current_price, na.rm = TRUE),
      mean(current_price, na.rm = TRUE),
      median(current_price, na.rm = TRUE),
      sd(current_price, na.rm = TRUE)
    )
  )

@@

In addition to the summary table, the distribution of current_price is visualized through a histogram (@fig-current_price), highlighting common price ranges and the overall spread. This visualization helps to identify any skewness or clustering in the data, which may indicate specific pricing patterns. Together, these tables and graphs provide a comprehensive understanding of the outcome variable, offering a detailed view of how product prices vary across different vendors and over time.

@@

The boxplot (@fig-current_price) presented above provides a detailed summary of the distribution of current_price for the products in the dataset. Boxplots are useful visualization tools for understanding the spread, central tendency, and overall variability of continuous data. In this case, the boxplot helps in identifying key features such as the median price, the interquartile range (IQR), and any potential outliers that may exist in the pricing data.

The median, represented by the horizontal line within the box, provides an indication of the central price of products in the dataset, highlighting where the bulk of prices are situated. The box represents the IQR, showing where the middle 50% of data points lie, thus giving insights into price concentration. The whiskers, extending from the box, represent the range within which most prices fall, excluding any extreme values or outliers. No outliers were visibly identified in this boxplot, which suggests that most product prices are consistently within a particular range.

The flat shape of the boxplot with a wide range of the IQR indicates low variability in current product prices. The relatively small difference between the lower and upper bounds of the box implies that the pricing of the products is consistent, with most prices concentrated around the median. This consistency may suggest that vendors are using similar pricing strategies or that there is limited differentiation in the types of products sold across vendors. Overall, the boxplot is a valuable tool for summarizing and communicating the underlying distribution of current_price, which aids in understanding the dynamics of pricing across different vendors.

## Predictor variables
The predictor variables in this study—month, old_price, and vendor—each play an important role in explaining variations in current_price. This section outlines how these variables potentially influence pricing, with a focus on capturing temporal patterns, historical effects, and vendor-specific pricing strategies. The summary tables and visual analyses help contextualize the impact of each predictor on the outcome, providing a broad understanding of the pricing dynamics within the dataset.

### Month
The month variable is used as a predictor to capture potential seasonal influences on pricing. By including month, we can observe if specific times of the year are associated with higher or lower prices. This is particularly important for identifying temporal patterns in pricing, which can be influenced by factors such as holidays, sales events, or seasonal demand changes. Table (@tbl-month_summary) provides a summary of the observations for each month, showing how frequently each month appears in the dataset. This information can help highlight any periods with higher or lower data collection frequency, which could influence the interpretation of seasonal pricing effects.

@@

### Old Price
old_price serves as another key predictor variable, representing the historical price of a product. This variable helps gauge the effect of historical pricing on current pricing strategies, indicating whether a product has undergone discounts or price hikes. Table (@tbl-old_price_summary) provides a summary of key statistics for old_price, including minimum, maximum, mean, median, and standard deviation. Understanding the distribution of old_price helps us assess whether historical prices significantly differ from current prices, and whether pricing adjustments (e.g., discounts) have been applied uniformly across products. The relationship between old_price and current_price is visually explored in Chart 4, which helps understand the influence of past pricing decisions on current prices.

@@

old_price_summary <- beef_data %>%
  summarise(
    Statistic = c("Min", "Max", "Mean", "Median", "Standard Deviation"),
    Value = c(
      Min = min(old_price, na.rm = TRUE),
      Max = max(old_price, na.rm = TRUE),
      Mean = mean(old_price, na.rm = TRUE),
      Median = median(old_price, na.rm = TRUE),
      `Standard Deviation` = sd(old_price, na.rm = TRUE)
  )
)

@@

### Vendor
The vendor variable is an important categorical predictor that differentiates pricing behavior across different retail chains. It allows us to analyze how pricing strategies vary among vendors such as Walmart and Galleria. Table (@tbl-vendor_summary) shows the count of observations for each vendor, allowing us to compare how frequently data was collected for each retail chain. This helps in understanding the representation of each vendor within the dataset and assessing whether certain vendors may have a stronger influence on the overall analysis. Chart 2, which shows the price difference by vendor, provides insights into how different vendors adjust their pricing strategies, offering valuable comparisons among them. Additionally, Chart 3 presents the average current_price over time, broken down by vendor, providing a clearer picture of how vendor-based strategies influence pricing trends.

@@

@@

To model the current price of beef, $y_i$, at time $i$, we use a Bayesian linear regression model implemented with the `stan_glm` function in the R package `rstanarm`. The response variable, $y_i$, represents the current price of beef in dollars. Our predictors include $x_1$, $x_2$, and $x_3$, which represent the month, old price, and vendor, respectively. Each component of the model is defined and justified below.

@@

Define $y_i$ as the current price of beef at time $i$, measured in dollars. Let $x_1$, $x_2$, and $x_3$ represent the predictors: $x_1$ is the month (categorical variable), $x_2$ is the old price, and $x_3$ is the vendor (categorical variable).

@@

In this model:
- $\alpha$ represents the intercept term, capturing the baseline price of beef.
- $\beta_1$ corresponds to the effect of the month on the current price, allowing us to capture seasonal variations that may impact beef pricing. The month is treated as a categorical variable, acknowledging that different months can bring different demand or supply effects due to holidays, weather, or other seasonal factors.
- $\beta_2$ represents the effect of the old price on the current price. This allows us to account for price inertia or trends where past pricing influences current pricing.
- $\beta_3$ represents the effect of the vendor, also treated as a categorical variable. Different vendors may have distinct pricing strategies, and including vendor as a categorical predictor helps us capture this vendor-specific pricing variation.

@@

These priors are chosen to be weakly informative, allowing the data to speak for itself while still providing a reasonable range for the coefficients based on prior expectations. Specifically, a standard deviation of 2.5 reflects our expectation that most reasonable effects should fall within a plausible range without being overly restrictive.

@@

The choice of predictors is well-balanced in terms of complexity and appropriateness for this scenario. By including month and vendor as categorical variables, we acknowledge the significant impact that both seasonal changes and vendor-specific factors can have on the current price of beef. The inclusion of the old price reflects the idea of price inertia, where historical prices are likely to influence current pricing trends.

The model is neither overly simplistic nor unnecessarily complex: it captures important predictors without adding extraneous complexity that could lead to overfitting. Month and vendor are treated as categorical variables rather than grouping by more arbitrary categories, which maintains the expressiveness of our model while being parsimonious.

The model is implemented using `stan_glm` from the `rstanarm` package, a high-level interface to Stan for Bayesian modeling. Stan provides powerful sampling algorithms, making it an appropriate tool for fitting our Bayesian model efficiently, ensuring robust convergence and accurate posterior estimation.

### Assumptions and Limitations
- **Linearity**: The model assumes a linear relationship between the predictors and the response variable. This may be a limitation if the true relationship is nonlinear.
- **Normality of Residuals**: The residuals are assumed to follow a normal distribution. If this assumption does not hold, it could bias the model predictions.
- **Priors**: We used weakly informative priors to allow flexibility. However, if we had prior knowledge of specific pricing patterns, more informative priors could be used for better inference.

### Validation and Diagnostics
- **Posterior Predictive Checks**: We performed posterior predictive checks to verify that the model adequately captures the variation in the data. These checks involve comparing simulated data from the posterior distribution to the observed data to check for any discrepancies.
- **Convergence**: Model convergence was assessed using trace plots and the R-hat statistic, with all values being close to 1, indicating successful convergence.
- **Alternative Models**: We also considered fitting simpler linear models that did not include the vendor or seasonal components. These models showed significantly lower predictive performance, indicating the importance of including both the vendor and seasonal factors for accuracy.

In summary, the chosen model provides a comprehensive yet interpretable understanding of how different factors influence the current price of beef. The Bayesian approach allows us to incorporate uncertainty and prior beliefs, while our specific choice of priors and predictors ensures that the model is well-suited for the available data and research questions.

@@

Figure 2 (@fig-current_price_distribution) presents the distribution of current beef prices in the dataset, offering insights into the range and frequency of beef pricing. The horizontal axis represents the price in dollars, ranging from 0 to approximately 25, while the vertical axis indicates the frequency, or the number of observations, within each price range. The grey bars illustrate the frequency distribution, showing how beef prices are spread across different price levels.

The histogram indicates that the most frequent price range is between \$5 and \$10, with a notable peak in the frequency around \$7 to \$8. This suggests that a significant proportion of beef products are priced within this range, making it a central pricing cluster in the market. Additionally, there are smaller peaks around \$13 and \$15, hinting at secondary groupings where beef prices are relatively common. The declining frequency beyond \$15 suggests fewer beef products are priced in the higher range, indicating that higher prices are less typical in this dataset.

The overlaid red line represents a smoothed density curve that helps to visualize the general shape and distribution trend of the prices. The curve follows the histogram closely, showing a skewed distribution with a concentration towards the lower prices, and tails off as prices increase. This pattern suggests that while lower-priced beef products are more common, higher prices are less frequent, potentially indicating a general preference or affordability constraint for lower-cost beef among consumers. The presence of multiple peaks could also imply different pricing tiers, perhaps influenced by factors such as beef quality or specific product types.

@@

Figure 3 (@fig-price_difference_by_vendor) illustrates the distribution of price differences between beef products offered by two vendors, T&T and Walmart, providing an insight into the variability of pricing across these retailers. The y-axis represents the price difference in dollars, while the x-axis labels the two vendors under comparison. The boxplots visually depict the range, median, and variability in the price differences for each vendor.

For T&T, the boxplot shows a wider range of price differences, with the interquartile range (IQR) extending from roughly \$1.5 to \$4, indicating a moderate to substantial variability in price differences among the products they sell. The median price difference is around \$2.5, suggesting that many of the products have a price difference near this value. However, T&T also shows significant variability with the presence of two outliers beyond \$7.5 and \$10, indicating that a few beef products exhibit much larger price differences compared to the rest of the data.

In contrast, Walmart exhibits a narrower spread of price differences, as indicated by the more compact boxplot. The IQR for Walmart lies approximately between \$1 and \$2.5, and the median price difference is lower compared to T&T, close to \$1.5. This suggests that Walmart’s beef products tend to have smaller and less variable price differences, indicating more consistent pricing practices. There are also a few outliers above \$4, but these are less extreme compared to T&T's outliers.

>>>>

The pricing of grocery items, particularly beef, plays an important role in understanding consumer behavior and market trends within the Canadian retail sector. As a staple in many households, beef pricing is shaped by various factors, including vendor strategies [@citePricingStrategies], historical trends, and seasonal variations [@citeSeasonal]. This study focuses on examining how major retailers adjust beef prices over time and how these adjustments reflect broader economic and cultural patterns. With growing concerns around food affordability and access, analyzing beef pricing dynamics is essential for both consumers and stakeholders involved in food distribution and policy-making.

This paper centers on two significant vendors, Walmart and T&T, to investigate how beef pricing differs across retail environments. Walmart, a global retailer, exemplifies standardized pricing strategies [@citeWalmart], while T&T, catering primarily to Asian communities, represents a more targeted market segment [@citeTandT]. By analyzing data from these vendors, the study addresses an often-overlooked aspect of how cultural preferences and retail strategies influence pricing. Previous research has focused on general trends in food pricing but has not fully considered the distinctions between mainstream retail chains and culturally specific stores. This analysis provides a detailed comparison of beef pricing patterns in these two diverse settings.

@@

Using a Bayesian linear regression model, the study examines how historical pricing, vendor-specific strategies, and seasonal factors impact current beef prices. The analysis indicates that past prices strongly predict current pricing, suggesting a high degree of consistency. Differences in pricing strategies between Walmart and T&T further highlight distinct market approaches, with Walmart focusing on stable pricing and T&T displaying greater variability, potentially to address specific consumer preferences. These results provide a clearer understanding of pricing strategies in a diverse retail landscape and offer insights into factors that shape affordability and market behavior.

The paper is structured as follows: @sec-data describes the data collection and cleaning methods, as well as the outcome and predictor variables used in the analysis. @sec-model introduces the forecasting models and explains their selection for predicting beef prices. @sec-result presents the key findings, including vendor-specific pricing effects and seasonal trends. Lastly, @sec-discussion interprets these results, comparing vendor strategies, identifying significant patterns, and discussing the study's limitations.

@@

The dataset for this research was obtained from Hammer, a publicly accessible repository offering pricing information from a variety of retail chains. It includes details on product attributes such as name, vendor, current and old prices, available units, and the month associated with certain prices. The dataset focuses on consumer pricing trends across different vendors, enabling an analysis of how factors like vendor type, historical pricing, and seasonal changes affect current prices. This provides a basis for examining retail market behaviors and assessing vendor pricing strategies.

The dataset contains both categorical and numerical variables. Important variables include ‘vendor,’ which identifies the retailer (e.g., Walmart, Galleria), and ‘product_name,’ which specifies the item being analyzed. Variables like ‘current_price’ and ‘old_price’ help track pricing changes, while temporal variables such as ‘month’ allow for the identification of seasonal pricing patterns. Summary statistics and visualizations have been generated to illustrate the distributions and relationships of these variables. These graphs include frequency distributions for vendors, trends in price changes, and comparisons between old and current prices. Relationships such as those between ‘vendor’ and ‘current_price’ are highlighted to present a detailed view of the dataset.

Alternative datasets, such as proprietary retail databases or other consumer purchasing records, were considered for this analysis. However, these were not selected due to limitations in accessibility, licensing restrictions, or insufficient detail in product-level attributes. The Hammer dataset was chosen because it provides detailed pricing data necessary for analyzing product-level trends across various vendors. It also enabled the creation of derived variables, such as ‘price_per_unit,’ which facilitated more meaningful comparisons between products. Data cleaning efforts included addressing missing values in the ‘old_price’ variable by replacing them with the ‘current_price,’ ensuring consistency and completeness in the analysis.

@@

The data for Project Hammer, focused on historical grocery prices, does not explicitly describe its collection methods. However, automated web scraping is a feasible approach for gathering this information. This method extracts structured information from retailer websites, capturing details such as product names, prices, brands, and unit measurements. The process involves developing scripts or utilizing tools that navigate websites, identify relevant HTML elements, and store the data in structured formats like CSV files or databases. To maintain accuracy and consistency, scraping schedules can be automated at regular intervals, such as weekly or monthly, to monitor price changes over time. Moreover, including user-agent strings and introducing delays in the scraping process helps simulate human browsing behavior and reduce the risk of triggering website anti-scraping mechanisms.

This research translates real-world phenomena into structured data entries within the dataset to analyze pricing behavior effectively. Variables such as ‘vendor,’ ‘current_price,’ and ‘month’ capture consumer purchasing patterns and their relationships with retail environments. The ‘vendor’ variable identifies the retailer, allowing an analysis of how retail settings influence pricing strategies. The ‘current_price’ and ‘old_price’ variables document pricing trends, enabling the measurement of changes in consumer costs over time and their connections to economic activity.

The ‘month’ variable introduces a temporal aspect, capturing the influence of seasonal patterns on pricing. Derived variables, such as ‘price_per_unit,’ standardize product comparisons by accounting for differences in package sizes or quantities. This transformation of consumer behavior and retail dynamics into structured data ensures that the analysis remains tied to real-world market activities, enhancing the understanding of pricing trends and behaviors across various retail contexts.

Additional details about the dataset are provided in the datasheet, which can be accessed through the repository associated with this paper.

## Outcome variables

The outcome variable for this study, current_price, is the central focus for analyzing product pricing patterns in the Canadian grocery market. It represents the price of beef at the time of data collection, providing a direct measure of how prices vary across different retail settings. By examining current_price, the study captures both short-term price changes and longer-term strategies used by vendors.

This section outlines the characteristics of the current_price variable, including its distribution and observed patterns, supported by summary statistics and visualizations. Key statistics, such as the mean, median, minimum, and maximum values, describe the central tendency and range of beef prices across vendors. Measures of variability, such as the standard deviation, illustrate the extent of price dispersion, showing how consistently prices are set across products and time. This descriptive analysis establishes a basis for identifying significant patterns in beef pricing.

### Current Price

This variable represents the price of the product at the time of data collection and serves as the foundation for analyzing how various factors influence pricing. By modeling current_price, the study examines the impact of historical pricing, vendor type, and seasonal patterns on current market prices. Understanding this variable is essential for identifying pricing strategies used by vendors and their effects on consumer expenses.

To present an overview of the outcome variable, a table is provided (@tbl-current_price). This table includes key statistics such as the minimum, maximum, mean, median, and standard deviation of the current_price variable. These measures summarize the distribution and variability of product prices, helping to characterize overall pricing patterns in the dataset.

@@

outcome_summary <- beef_data %>%
  summarise(
    Statistic = c("Min", "Max", "Mean", "Median", "Standard Deviation"),
    Value = c(
      round(min(current_price, na.rm = TRUE), 3),
      round(max(current_price, na.rm = TRUE), 3),
      round(mean(current_price, na.rm = TRUE), 3),
      round(median(current_price, na.rm = TRUE), 3),
      round(sd(current_price, na.rm = TRUE), 3)
    )
  )

@@

@tbl-current_price highlighting key measures of central tendency and variability. The data ranges from a minimum value of 0.77 to a maximum of 23.98, illustrating a broad spread. The mean value is 9.11, indicating that most beef prices are concentrated around this range. The median is 8.58, suggesting that half of the observations fall below this value. The proximity of the mean and median implies a relatively symmetric distribution. The standard deviation of 4.33 reflects moderate variability, indicating that most values are distributed within approximately 4.33 units of the mean, providing a clearer understanding of the data's spread.

In addition to the summary table, @fig-current_price visualizes the distribution of current_price, showing common price ranges and the overall spread. This graph helps identify any skewness or clustering in the data, which could indicate specific pricing patterns. Combined, the table and histogram provide a detailed overview of the outcome variable, illustrating how product prices differ among vendors and change over time.

@@

The @fig-current_price provides an overview of the distribution of current_price for the products in the dataset. This visualization highlights key features such as the median price, interquartile range (IQR), and any potential outliers in the pricing data.

The median, represented by the horizontal line within the box, indicates the central tendency of product prices, showing where most values are centered. The box itself represents the IQR, which captures the range within which the middle 50% of prices fall, offering a view of price concentration. The whiskers extending from the box depict the range of prices outside the IQR, excluding extreme values. In this case, no significant outliers were detected, suggesting that product prices are generally within a consistent range.

The relatively flat boxplot, combined with a narrow IQR, indicates limited variability in current_price. The small difference between the lower and upper bounds of the box suggests that product pricing is consistent, with most prices clustering around the median. This consistency could reflect standardized pricing strategies among vendors or a lack of variation in the types of products sold. Overall, the boxplot effectively summarizes the distribution of current_price and provides a clear view of pricing patterns across vendors.

## Predictor variables

The predictor variables in this study—month, old_price, and vendor—each contribute to explaining variations in current_price and provide a structured view of the factors influencing beef pricing across retail settings. These variables reflect temporal trends, the influence of historical pricing, and differences in vendor strategies, collectively aiding in the understanding of product pricing in the Canadian grocery market.

The month variable represents the temporal aspect of pricing, enabling the analysis of seasonal patterns in beef prices. Treating month as a categorical variable allows for the identification of months with higher or lower prices, potentially tied to changes in demand due to holidays, promotions, or seasonal supply fluctuations. For instance, months associated with holidays like Thanksgiving or cultural festivals may show increased demand, which could lead to price adjustments. This temporal perspective helps to assess whether vendors apply seasonal pricing strategies that influence current_price over time.

### Month

The month variable serves as a predictor to account for seasonal effects on pricing. Including month allows the analysis to identify whether certain times of the year are associated with price fluctuations. Factors such as holidays, promotional events, or changes in seasonal demand may contribute to these variations. @tbl-month_summary presents a breakdown of the observations for each month, showing the frequency of data collection across the year. This summary helps identify any months with disproportionately high or low representation, which could impact the interpretation of seasonal pricing patterns.

@@

The @tbl-month_summary presents the number of observations recorded for each month, spanning June to November. The highest count is in August (624), while the lowest is in November (342). This variation may reflect increased market activity or data collection efforts during the summer, possibly due to higher consumer demand during barbecue season. Conversely, the decline in November could be linked to reduced data collection or changes in consumer purchasing patterns. Since data collection began in early 2024, the dataset does not represent the entire year, limiting the ability to evaluate pricing trends across all seasons. These monthly differences highlight the importance of accounting for seasonal factors when analyzing pricing patterns.

### Old Price

old_price is a key predictor variable that represents a product's historical price. This variable is used to examine how past pricing influences current pricing strategies, indicating whether products have experienced discounts or price increases. @tbl-old_price_summary summarizes the key statistics for old_price, including its minimum, maximum, mean, median, and standard deviation. Analyzing the distribution of old_price allows us to evaluate whether historical prices significantly differ from current prices and whether adjustments, such as discounts, are applied consistently across products. The relationship between old_price and current_price is further illustrated in Chart 4, providing a visual representation of how historical pricing affects current price decisions.

@@

old_price_summary <- beef_data %>%
  summarise(
    Statistic = c("Min", "Max", "Mean", "Median", "Standard Deviation"),
    Value = c(
      round(min(current_price, na.rm = TRUE), 3),
      round(max(current_price, na.rm = TRUE), 3),
      round(mean(current_price, na.rm = TRUE), 3),
      round(median(current_price, na.rm = TRUE), 3),
      round(sd(current_price, na.rm = TRUE), 3)
  )
)

@@

The @tbl-old_price_summary summarizes the statistics for the predictor variable old_price, representing the historical price of beef. The minimum value is \$1.87, and the maximum value is \$25.99, reflecting a wide range of historical prices for beef products. The mean value of \$11.48 indicates that, on average, beef products were priced near this level in the past. The median of \$10.97 shows that half of the products had historical prices below this value, suggesting a slightly skewed distribution. The standard deviation of \$5.32 reflects moderate variability, indicating that while many products were priced close to the average, some were significantly higher or lower. These statistics help describe the variation and central tendency in historical beef pricing.

### Vendor

The vendor variable is a categorical predictor that captures differences in pricing behavior across retail chains. It enables the analysis of how pricing strategies differ among vendors. @tbl-vendor_summary lists the number of observations for each vendor, helping to evaluate the representation of each retail chain in the dataset and determine whether certain vendors have a larger impact on the overall analysis.

@@

The @tbl-vendor_summary presents the number of observations for each vendor, with Walmart accounting for 1,696 and T&T for 1,330. While there is a difference in the number of observations, the counts are close enough to allow for a meaningful comparison between the two vendors. This indicates that while Walmart may offer a slightly wider range or maintain more consistent availability of beef products, T&T’s dataset is still robust enough to analyze pricing patterns effectively. This balance ensures the analysis reflects variations between a mainstream retailer and a culturally focused store without being significantly affected by unequal sample sizes.

@@

To model the current price of beef, $y_i$, at time $i$, we apply a Bayesian linear regression model using the `stan_glm` function from the R package `rstanarm`. The response variable, $y_i$, represents the current price of beef in dollars. The predictors include $x_1$, $x_2$, and $x_3$, corresponding to the month, old price, and vendor, respectively. Each component of the model is defined and explained in the following sections.

@@

Define $y_i$ as the current price of beef at time $i$, expressed in dollars. The predictors are represented as follows: $x_1$ is the month (a categorical variable), $x_2$ is the old price, and $x_3$ is the vendor (a categorical variable).

@@

In this model:
- $\alpha$ denotes the intercept term, representing the baseline price of beef when all predictors are at their reference levels.  
- $\beta_1$ represents the effect of the month on the current price, capturing seasonal patterns that may influence beef prices. The month is treated as a categorical variable, recognizing that demand and supply conditions can vary by month due to factors such as holidays, weather, or promotional periods.  
- $\beta_2$ reflects the effect of the old price on the current price, accounting for the influence of historical pricing. This term helps identify whether past prices have a consistent impact on current pricing decisions.  
- $\beta_3$ captures the effect of the vendor, also treated as a categorical variable. Different vendors may apply unique pricing strategies, and including this variable allows us to model variations specific to each vendor.

@@

These priors are designed to be weakly informative, offering guidance while allowing the data to determine the outcomes. A standard deviation of 2.5 reflects the expectation that most effects are likely to fall within a reasonable range, ensuring the priors are flexible without imposing overly restrictive assumptions.

@@

The selected predictors strike a balance between relevance and simplicity for this scenario. Including `month` and `vendor` as categorical variables accounts for the influence of seasonal changes and vendor-specific pricing strategies on the current price of beef. The use of `old_price` reflects the concept of price inertia, where past prices are likely to affect current pricing patterns.

The model avoids being overly simplistic or unnecessarily complex by incorporating key predictors without introducing excessive variables that could risk overfitting. Treating `month` and `vendor` as categorical variables ensures the model captures relevant distinctions without arbitrary groupings, maintaining both clarity and parsimony.

The model is implemented using `stan_glm` from the `rstanarm` package, which provides an accessible interface to Stan for Bayesian modeling. Stan's efficient sampling algorithms ensure reliable convergence and accurate estimation of posterior distributions, making it a robust choice for this analysis.

### Assumptions and Limitations
- **Linearity**: The model assumes a linear relationship between the predictors (month, old_price, vendor) and the response variable (current_price). This assumption simplifies the analysis and enhances interpretability, but it may limit the model's ability to capture more complex relationships. For example, interactions or threshold effects between predictors and the response variable would not be reflected in a purely linear framework. If these nonlinear relationships exist but are not modeled, the resulting estimates and predictions may be less accurate, potentially leading to incorrect conclusions about factors affecting beef pricing.

- **Normality of Residuals**: The residuals are assumed to follow a normal distribution, a requirement for making valid statistical inferences about model parameters. If this assumption is violated—such as when residuals are skewed or have heavy tails—it could result in biased estimates, unreliable confidence intervals, and less dependable predictions. Non-normal residuals might suggest missing variables in the model or that a different modeling approach, such as a generalized linear model, might better fit the data.

- **Priors**: The Bayesian framework employs weakly informative priors to allow the data to primarily determine parameter estimates. While this provides flexibility, it also means that stronger prior knowledge of pricing patterns could have been used to improve precision. Using more specific priors, informed by domain knowledge, might enhance the model's performance by producing more stable estimates. However, weakly informative priors can make the model more sensitive to data quality and outliers, as they do not strongly influence the parameter estimates. In cases where the data are limited or noisy, stronger priors could help stabilize the model's outputs.

### Validation and Diagnostics
- **Posterior Predictive Checks**: Posterior predictive checks were conducted to evaluate how well the model represents the variation in the data. These checks compare simulated data generated from the posterior distribution to the observed data to identify any discrepancies. Assessing the alignment between these distributions helps determine whether the model captures the underlying patterns in the data. A close match indicates that the model is appropriately capturing the structure of the data, while notable differences may suggest issues such as omitted variables or an incorrect model specification. These checks are a key step in validating the model's ability to generalize beyond the observed dataset.

- **Convergence**: Model convergence was assessed using trace plots and the R-hat statistic, with all R-hat values near 1, confirming successful convergence. Trace plots illustrate how effectively the Markov Chain Monte Carlo (MCMC) sampling algorithm explores the parameter space. Well-mixed chains with stable trajectories suggest the model has reached a stationary distribution. The R-hat statistic, also known as the Gelman-Rubin diagnostic, compares between-chain and within-chain variance, with values close to 1 indicating that all chains have converged to the same target distribution. Together, these diagnostics ensure the reliability of the parameter estimates and confirm that sampling issues did not influence the results.

- **Alternative Models**: Simpler linear models excluding vendor or seasonal variables were also tested. These models showed notably lower predictive performance, highlighting the importance of including both vendor and seasonal factors for accurate modeling. Excluding these variables reduced the model's fit, as indicated by metrics like R-squared and predictive accuracy. This comparison reinforces the necessity of accounting for vendor-specific strategies and seasonal variations to effectively model beef pricing. The evaluation of alternative models confirmed that the selected predictors provide a better representation of the factors influencing price changes in the dataset.

In conclusion, the chosen model effectively captures how various factors influence the current price of beef. The Bayesian framework incorporates uncertainty and prior knowledge, while the selection of priors and predictors ensures the model aligns well with the available data and research objectives.

@@

@fig-current_price_distribution illustrates the distribution of current beef prices in the dataset, depicting how prices are spread across different levels. The horizontal axis represents beef prices in dollars, ranging from 0 to approximately 25, while the vertical axis shows the frequency, or the number of observations, within each price range. The grey bars represent the histogram, visualizing the frequency distribution of beef prices.

The histogram shows that the most common price range is between \$4 and \$8, with noticeable peaks around \$5 and \$8. This indicates that a significant portion of beef products falls within this range, suggesting a central pricing cluster in the market. Smaller peaks around \$13 and \$15 point to secondary groupings where beef prices are also relatively frequent. Beyond \$15, the frequency declines sharply, indicating that higher-priced beef products are less common and that most items are concentrated within a lower price range.

The overlaid red line represents a smoothed density curve, highlighting the overall trend in the distribution. The curve closely aligns with the histogram, showing a skewed pattern with a higher concentration of prices in the lower range and a gradual decrease as prices rise. This distribution suggests that consumers may prefer lower-priced beef products or that pricing aligns with affordability constraints. The multiple peaks in the curve indicate distinct pricing tiers, potentially reflecting differences in product quality, cut types, or retailer-specific pricing strategies.

@@

@fig-price_difference_by_vendor shows the distribution of price differences for beef products sold by T&T and Walmart, highlighting variations in pricing between these retailers. The y-axis represents the price difference in dollars, while the x-axis lists the two vendors. The boxplots display the range, median, and variability of price differences for each retailer.

For T&T, the boxplot indicates a broader range of price differences, with the interquartile range (IQR) spanning approximately \$1.5 to \$4. The median price difference is about \$2.5, suggesting that many products cluster around this value. T&T also shows greater variability, with outliers extending beyond \$7.5 and \$10. These outliers suggest that certain products have significantly larger price differences, possibly reflecting diverse pricing strategies or variations in product offerings.

In comparison, Walmart's boxplot shows a narrower distribution of price differences. The IQR ranges from roughly \$1 to \$2.5, with a median of around \$1.5. This tighter range indicates that Walmart’s beef pricing differences are more consistent, reflecting a standardized pricing strategy. While there are a few outliers above \$4, these deviations are less pronounced than those seen for T&T.
